---
title: "Week4 Work"
format: html
editor: visual
---

## Exploration 3 week 4

## 

```{r}
library(tidyverse)

gapminder <- read_csv(
  "https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv")

# gapminder <- read_csv("gapminder-FiveYearData.csv")  # if you saved the CSV in the same folder as this qmd
```

-   **Wealth, Health, and Tension in the Data**

    **Guiding question:\
    **To what extent does economic wealth explain differences in life expectancy, and where does this explanation break down? \This is not a yes/no question. You are expected to explore **patterns and tensions** in the data.

This graph shows continents by color comparing GDP per capita and Life Expectancy
```{r}
gapminder |>
  group_by(continent) |>
  summarize(avg_lifeExp = mean(lifeExp), avg_gdp = mean(gdpPercap))
gapminder |>
  filter(year == 2007) |>
  ggplot(aes(x = lifeExp, y = gdpPercap, color = continent)) + 
  geom_point() +
  scale_x_log10() +
  labs(
    title = "GDP per Capita & Life Expectancy Comparison",
    x = "GDP",
    y = "Life Expectancy"
  ) +
  theme_minimal()
This graph shows countries by color comparing GDP per capita and Life Expectancy
{r}
gapminder |>
  group_by(continent) |>
  summarize(avg_lifeExp = mean(lifeExp), avg_gdp = mean(gdpPercap))
gapminder |>
  filter(continent == "Americas") |>
  ggplot(aes(x = lifeExp, y = gdpPercap, color = country)) + 
  geom_point() +
  scale_x_log10() +
  labs(
    title = "GDP per Capita & Life Expectancy Comparison(Countries)",
    x = "GDP",
    y = "Life Expectancy"
  ) +
  theme_minimal()
```



