---
title: "Week4 Work"
format: html
editor: visual
---

## Exploration 3 week 4

## 

```{r}
library(tidyverse)

gapminder <- read_csv(
  "https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv")

# gapminder <- read_csv("gapminder-FiveYearData.csv")  # if you saved the CSV in the same folder as this qmd
```

-   **Wealth, Health, and Tension in the Data**

    **Guiding question:\
    **To what extent does economic wealth explain differences in life expectancy, and where does this explanation break down? This is not a yes/no question. You are expected to explore **patterns and tensions** in the data.

    ### **Ways to sharpen the question (examples)**

    You might choose to:

-   Compare the GDP–life expectancy relationship across continents

-   Examine how this relationship changes over time

-   Identify years or regions where the relationship appears weaker or stronger

Example analytical artifacts:

-   A table of average GDP per capita and life expectancy by continent and year

-   Scatterplots that make patterns and deviations visible

### **Planning prompt**

Explain:

-   What would it mean, in data terms, for the relationship to be “strong” or “break down”?

-   How will your visualizations help you detect this?

```{r}
gapminder |>
  group_by(continent) |>
  summarize(avg_lifeExp = mean(lifeExp), avg_gdp = mean(gdpPercap))
gapminder |>
  filter(year == 2007) |>
  ggplot(aes(x = lifeExp, y = gdpPercap, color = continent)) + 
  geom_point() +
  scale_x_log10() +
  labs(
    title = "GDP per Capita & Life Expectancy Comparison",
    x = "GDP",
    y = "Life Expectancy"
  ) +
  theme_minimal()

```
